Nonparametric Clustering from Energy Statistics

### 2017-09-17 qcqp_clustering/draft/energy_clustering_v2.tex

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
% @gui: not sure what you mean?  what counts as machine learning?  be specific, i think yo mean clustering?  even then, it is not true, both the master's thesis and the RKHS paper kind of do it.  be as general as possible about the gap that we are filling, but not more general.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else?  
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models in a variety of settings.
\end{abstract}

@@

The clustering problem has a long history in machine learning.
Perhaps the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and it is thus sensitive to outliers. 
% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator. 
Moreover, $k$-means
is strongly parametric, assuming that data is isotropically and normally
distributed since it is a special case of gaussian mixture models (GMM) 
% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.  k-means is certainly closely related to GMM though.
through the expectation maximization (EM) algorithm. Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space, with a clear resolution of cluster centers and without
significant overlap of the points belonging to different clusters.
% i@gui:  don't know what "clear resolution of cluster centers" adds to this?

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. This enables one to cluster way more
complicated datasets, 
% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
however, the kernel choice remains the biggest challenge
since there is no principled theory behind this, 
% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
and it usually introduces
extra parameters that need to be optimized. 
% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement, showing that there is only one possible optimization problem
consistent with energy statistics. 
% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
Moreover, we show that this is
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
\cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local 
solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}.
In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a family of 
default kernels, showing that
it is able to perform accurately on datasets coming very different 
distributions, contrary to $k$-means and GMM for instance.
% @gui: to me, we did not make explicit one of the most important points.   in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically).

% @gui: i find ToC paragraphs useless.
Our work is organized as follows. In section~\ref{sec:background} we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard. We also show the equivalence to 
kernel $k$-means problem.
In Section~\ref{sec:weighted} we generalize these results to a weighted
version of energy statistics, which provides the connection with graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm based on Hartigan's method
to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

% @gui: let the reader decide what is important, rather than us telling them :)
It is important to 
mention that \eqref{eq:energy2} can be even further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legitamate metric.

@@

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$. Moreover, this can be estimated from a sample
according to \eqref{eq:inner_data}.
% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric test statistic} for equality of distributions
\cite{Szkely2013}.
% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
Due to the test statistic \eqref{eq:between} for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, $S$ should attain a maximum when each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing \eqref{eq:between} is, however, equivalent to minimizing
\eqref{eq:within} which has a more convenient form.
% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).

\begin{proposition}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type \eqref{eq:negative_type}. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes \eqref{eq:between} if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k),
\end{equation}
where $W$ is given by \eqref{eq:within}.
\end{proposition}
\begin{proof}
From \eqref{eq:within} and \eqref{eq:between}
we have
\begin{equation}
\begin{split}
S + W &= 
\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
+ \dfrac{1}{2n} \sum_{i=1}^{k} 
\bigg[ n - 
\sum_{j\ne i = 1}^k n_j \bigg] 
n_i g(\C_i, \C_i) 
\\
& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)
= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
\end{split}
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ over the choice of partition is equivalent to minimizing $W$.
\end{proof}
% @gui: do other people know this result? if so, cite them, if not, state that.

For a given $k$, the clustering problem amounts to
finding the best partition of the data by solving~\eqref{eq:minimize}.
% @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
Notice that this is a hard clustering problem as partitions
are disjoint. The problem \eqref{eq:minimize} based on
energy statistics was already proposed in \cite{Kgroups}. However, it is
important to note that this is equivalent to maximizing \eqref{eq:between}
which is the test statistic for equality of distributions. In the following
we show what is the explicit optimization problem behind \eqref{eq:minimize}.

% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
Let us formulate \eqref{eq:minimize} in the corresponding
RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Let us define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
\vdots & \vdots & \ddots  & \vdots \\
\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$ where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. Let us also introduce the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP).

>>>>

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces 
was proposed.
%Although  
%extensively used in statistics, it
%has not been previously incorporated in machine learning problems.
% @gui: not sure what you mean?  what counts as machine learning?  be specific, i think yo mean clustering?  even then, it is not true, both the master's thesis and the RKHS paper kind of do it.  be as general as possible about the gap that we are filling, but not more general
% @jovo: I think we should just remove the phrase. 
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
% @gui: have other people shown a relationship between kernel k-means and say, RKHS, or something else? 
% @jovo: yes, any kernel method is defined on a HKHS. 
% For instance ref [11] is the first one to consider kernel k-means in a 
% RKHS. The whole problem is how do you formulate a 
% statistical problem in a RKHS. In the case of
% energy this was done only by [6], and it's based on this work that we 
% have the theory presented in this paper
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
quality. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

The clustering problem has a long history in machine learning.
Perhaps the most used method is $k$-means \cite{Lloyd,MacQueen,Forgy}, which
is based on Lloyd's heuristic \cite{Lloyd} of assigning a data point to
the cluster with closest center. The only statistical 
information about each cluster
comes from its mean and it is thus sensitive to outliers.
% @gui: algorithms don't make assumptions, theorems do.  also, algorithms cannot be parametric or non-parametric, that is a property only of the statistical model, or an estimator. 
%Moreover, $k$-means
%is strongly parametric, assuming that data is isotropically and normally
%distributed since it is a special case of gaussian mixture models (GMM) 
% @gui: k-means is NOT a special case of GMM, GMM is a soft clustering (not fuzzy, in stats and machine learning we don't say "fuzzy"), where as k-means is a hard clustering.  k-means is certainly closely related to GMM though.
%through the expectation maximization (EM) algorithm. 
Nevertheless,
$k$-means works very well when data is linearly separable 
in Euclidean space.
%, with a clear resolution of cluster centers and without
%significant overlap of the points belonging to different clusters.
% @gui:  don't know what "clear resolution of cluster centers" adds to this?
% @jovo: commented out some parts above and included new phrase here.
Gaussian mixture models (GMM) is also very commonly used for 
clustering,
however it is strongly parametric, as $k$-means which is closely related.

To account for nonlinearities, kernel methods were introduced 
\cite{Smola,Girolami}. A mercer kernel \cite{Mercer} is used to implicitly
map data points to a RKHS, then clustering can be performed in the associated
Hilbert
space by using its inner product. 
%This enables one to cluster way more complicated datasets, 
This enables one to cluster data that is not separable in Euclidean space.
% @gui: algorithms cluster, doesn't matter how complicated the data are.  so, what do you actually mean here?  perform well empirically? what does "complicated" actually mean?
% @jovo: what I mean is not tied to any algorithm. I want to say that kernel
% methods allows one to cluster data that is not separable in Euclidean space.
% I modified the phrase, see if it's better
However, the kernel choice remains the biggest challenge
since there is no principled theory to construct a kernel for a given dataset,
%behind this, 
% @gui: the words "this" and "it" are often vague and ambiguous, be specific/precise
and usually a kernel introduces
hyperparameters that need to be carefully chosen.
% @gui: not sure what you mean here, in kernel k-means, we don't "optimize" over any parameters typically.  sometimes we select hyperparameters of kernels via cross-validation, is that what you mean?
% @jovo: yes this is what I mean, tried to fix it
The well-known
kernel $k$-means optimization problem is nothing but $k$-means in the
feature space \cite{Girolami}. Furthermore, kernel $k$-means algorithm
\cite{Dhillon2,Dhillon} is still based on Loyd's heuristic \cite{Lloyd} of
using the mean of each cluster in the feature space. 
We refer the reader to \cite{Filippone} for a survey of clustering
methods.

Although clustering from energy statistics was considered
in \cite{Kgroups}, the precise optimization problem behind this approach
remains elusive, as well as the connection to kernel methods.
The main theoretical contribution of this paper is to fill this gap.
Since the statistical potential energy is minimum when
distributions are equal, the
principle behind clustering is to maximize the statistical energy, 
enforcing probability distributions associated to each cluster
to be different from one another.
We provide a precise mathematical formulation to this
statement
%, showing that there is only one possible optimization problem
%consistent with energy statistics. 
% @gui: not sure what you mean "only one possible optimization problem".  we can reformulate it many different ways? also, below, "this" is ambiguous.
% @jovo: I had in mind that minimizing W is equivalent to maximizing S, 
% but it's confusing at this point so just removed the phrase
leading to
a quadratically constrained
quadratic program (QCQP) in the associated RKHS. Our results immediately 
establish the connection 
to kernel methods commonly used in machine learning, 
by showing that this QCQP is equivalent
to kernel $k$-means optimization problem. 
The equivalence between kernel $k$-means, 
spectral clustering, and graph partitioning problems was already established
\cite{Dhillon,Dhillon2}. We show how these connections arise
from a weighted version of energy statistics.

Our algorithmic contribution is to use Hartigan's method
\cite{Hartigan} to find local 
solutions of the above mentioned QCQP.
This approach was also proposed in \cite{Kgroups}.
In this case,
the resulting algorithm has the same
time complexity as kernel $k$-means algorithm, however, the numerical 
results provide compelling evidence that this method
is more accurate and robust than kernel $k$-means. This evidence
is also supported by general theoretical results \cite{Telgarsky,Slonin}.
Moreover, in our experiments we put in
evidence the nonparametric aspect of energy
statistics based clustering, which provides a family of 
default kernels, showing that
it is able to perform accurately on datasets coming from very different 
distributions, contrary to $k$-means and GMM for instance.
% @gui: to me, we did not make explicit one of the most important points.   in this paper, we make clear exactly when we expect energy clustering to perform better than k-means/GMM, and when we expect it to perform worse (empirically). 
% @jovo: added this phrase, check it
More specifically, the proposed method performs closely to $k$-means
and GMM on normally distributed data with balanced clusters, however, it
performs considerably better on  data that is not normally distributed. 
It also
performs better than $k$-means and GMM in higher dimensions, even on 
gaussian settings, and it performs worse than 
GMM when we have highly unbalanced clusters.

% @gui: i find ToC paragraphs useless.
% @jovo: I agree and don't mind removing, however it's a tradition that 
% most referees like to see, and it doesn't do any harm. I'm fine with
% whatever choice
Our work is organized as follows. In section~\ref{sec:background} we introduce
the necessary background on energy statistics and RKHS.
Section~\ref{sec:clustering_theory} contains the main theoretical 
results of this paper,
where we consider a clustering theory based on energy statistics leading
to a QCQP, which is NP-hard. We also show the equivalence to 
kernel $k$-means problem.
In Section~\ref{sec:weighted} we generalize these results to a weighted
version of energy statistics, which provides the connection with graph
partitioning problems and spectral clustering.
In Section~\ref{sec:twoclass} we consider a simple example in one dimension,
where we propose an algorithm which requires no initialization.
In section~\ref{sec:algo} we briefly review kernel $k$-means algorithm,
and propose a new iterative algorithm based on Hartigan's method
to solve this QCQP.
Section~\ref{sec:numerics} contains some carefully designed numerical
experiments indicating that this algorithm outperforms kernel
$k$-means, standard $k$-means, and GMM/EM algorithms.
Our final conclusions are presented in section~\ref{sec:conclusion}.

@@

% @gui: let the reader decide what is important, rather than us telling them :)
%It is important to mention 
The energy distance \eqref{eq:energy2} can be further generalized.
Let $X, Y \in \mathcal{X}$  where $\mathcal{X}$ is an arbitrary space endowed
with a \emph{semimetric of negative type}
$\rho: \mathcal{X}\times\mathcal{X} \to \mathbb{R}$, which is required
to satisfy
\begin{equation}
\label{eq:negative_type}
\sum_{i,j=1}^n c_i c_j \rho(X_i, X_j) \le 0,
\end{equation}
where $X_i \in \mathcal{X}$ and $c_i \in \mathbb{R}$ such that
$\sum_{i=1}^n c_i = 0$. Then, $\mathcal{X}$ is called a \emph{space of
negative type}.
We can thus replace $\mathbb{R}^D \to \mathcal{X}$ and 
$\| X - Y \| \to \rho(X , Y)$ in the definition \eqref{eq:energy}, obtaining
the generalized energy distance
\begin{equation}
\label{eq:energy3}
\Energy(P, Q) \equiv 2 \E \rho(X,Y) - \E \rho(X, X') - \E \rho(Y,Y').
\end{equation}
For spaces of negative type there exists a Hilbert space $\mathcal{H}$ and
a map $\varphi: \mathcal{X} \to
\mathcal{H}$ such that
$\rho(X, Y) = \| \varphi(X) - \varphi(Y) \|_{\mathcal{H}}^2$. This
allows us to compute quantities related to probability distributions over
$\mathcal{X}$ in the Hilbert space $\mathcal{H}$.
Even though the semimetric 
$\rho$ may not satisfy the triangle inequality, 
$\rho^{1/2}$ does since it can be shown to be a legitamate metric.

@@

Now we can state the equivalence between energy distance $\Energy$ and
inner products on RKHS, which is one of the main results of
\cite{Sejdinovic2013}. If $\rho$ is a semimetric
of negative type and $\kk$ a kernel that generates $\rho$, then
replacing \eqref{eq:gen_kernel} into
\eqref{eq:energy3}, and using \eqref{eq:mmd2}, yields
\begin{equation} \label{eq:Erho}
\Energy(P, Q) = 
2 \left[ \E \, \kk(X, X') + \E \, \kk(Y, Y') - 2\E \, \kk(X, Y)\right] 
= 2 \gamma_\kk^2(P,Q) .
\end{equation}
Due to \eqref{eq:mmd} we can compute the energy distance using the inner 
product of $\Hk$. 
%Moreover, this can be estimated from a sample
%according to \eqref{eq:inner_data}.
% @gui: i don't know what this means.  eqn 7 defines something for a population statistic, not a sample statistic.
% @jovo: removed. I was thinking that you can always estimate the expected 
% value by computing a sample mean. but I agree, it's confusing

Finally, let us recall the main formulas from energy statistics
for the test statistic of equality of distributions \cite{Szkely2013}. 
Assume we have data $\mathbb{X} = \{ x_1,\dotsc, x_n \}$ where
$x_i \in \mathcal{X}$, and $\mathcal{X}$ is a space of negative type.
Consider a disjoint partition $\mathbb{X} = \bigcup_{j=1}^k \C_j$, with
$\C_i \cap \C_j = \emptyset$.
Each expectation in 
\eqref{eq:energy3}
can be computed 
through the function
\begin{equation}
\label{eq:g_def}
g (\C_i, \C_j) \equiv 
\dfrac{1}{n_i n_j}
\sum_{x \in \C_i} 
\sum_{y \in \C_j} \rho(x, y)
\end{equation}
where $n_i = |\C_i|$ is the number of elements in 
$\C_i$. 
The \emph{within energy dispersion} is defined by
\begin{equation}
\label{eq:within}
W \equiv
\sum_{j=1}^{k} \dfrac{n_j}{2} g(\C_j, \C_j)
\end{equation}
and the \emph{between-sample energy statistic} is defined by
\begin{equation}
\label{eq:between}
S \equiv
\sum_{1 \le  i < j \le k } \dfrac{n_i n_{j}}{2 n} \left[
2 g(\C_i, \C_j) - 
g(\C_i, \C_i) - 
g(\C_j, \C_j)
\right]
\end{equation}
where $n = \sum_{j=1}^k n_j$.
Given a set of distributions
$\{ P_j\}_{j=1}^k$, where $x \in \C_j$ if and only if $x \sim P_j$, 
the quantity $S$ provides
a \emph{nonparametric test statistic} for equality of distributions
\cite{Szkely2013}.
% @gui: i'm not sure the above is true/meaningful.  nonparametric means implicitly estimating an infinite number of parameters.  without any statistical theory, i'm not quite sure what we can say.
% @jovo: my understanding is that this hypothesis test does not assume any
% form of the distribution. For instance, Kolmogorov-Smirnov is also
% said to be nonparametric
When the sample size is large enough, $n\to \infty$,
under the null hypothesis $H_0: P_1=P_2=\dotsm=P_k$ we have that
$S\to 0$, 
and under
the alternative hypothesis $H_1: P_i \ne P_j$ for at least two $i\ne j$, 
we have that $S \to \infty$.
This test is nonparametric in the sense that it does not make any assumptions
about the distributions $P_j$.

@@

% @gui: consider only using equation numbers for equations that you later refer to? there are so many, it is easy to get lost.
% @jovo: ok will do that after we are done revising the paper. This is a matter
% of taste, some people enumerate all equations, others just the ones cited
Due to the test statistic \eqref{eq:between} for equality of distributions,
the obvious
criterion for clustering data is to 
maximize $S$ which makes 
each cluster as different
as possible from the other ones.
In other words, given a set of points coming from different probability
distributions, $S$ should attain a maximum when each point is correctly
classified as belonging to the cluster associated to its probability
distribution.
The following 
straightforward result
shows that maximizing $S$ is, however, equivalent to minimizing
$W$ which has a more convenient form.
% @gui: where possible, i recommend using the "name" of the quantity rather than the number of the equation.  eg, here, write S and W, rather than (13) and (12).
% @jovo: i think this is fine for the main symbols, will do

\begin{proposition}
\label{th:minimize}
Let $\mathbb{X} = \{x_1,\dotsc,x_n\}$ where each data point
$x_i$ lives in a space $\mathcal{X}$ endowed with a semimetric $\rho:
\mathcal{X}\times\mathcal{X} \to \mathbb{R}$ of
negative type \eqref{eq:negative_type}. For a fixed integer $k$,
the partition
$\mathbb{X} = \bigcup_{j=1}^k \C_j$, where $\C_i \cap C_j = \emptyset$ for
all $i\ne j$, maximizes $S$ if and only if
\begin{equation}
\label{eq:minimize}
\min_{\C_1,\dotsc,C_k  } W(
\C_1, \dotsc, \C_k).
\end{equation}
\end{proposition}
\begin{proof}
From \eqref{eq:within} and \eqref{eq:between}
we have
\begin{equation}
\begin{split}
S + W &= 
\dfrac{1}{2n} \sum_{\substack{i,j=1 \\ i\ne j}}^k n_i n_j g(\C_i, \C_j)
+ \dfrac{1}{2n} \sum_{i=1}^{k} 
\bigg[ n - 
\sum_{j\ne i = 1}^k n_j \bigg] 
n_i g(\C_i, \C_i) 
\\
& = \dfrac{1}{2n} \sum_{i,j=1}^k n_i n_j g(\C_i, \C_j)
= \dfrac{1}{2n} \sum_{x \in \mathbb{X}} \sum_{y \in \mathbb{X}} \rho(x,y)
= \dfrac{n}{2} g(\mathbb{X}, \mathbb{X}).
\end{split}
\end{equation}
Note that the right hand side of this equation 
only depends on the pooled data, so it is a constant
independent of the choice of partition. Therefore, maximizing
$S$ over the choice of partition is equivalent to minimizing $W$.
\end{proof}
% @gui: do other people know this result? if so, cite them, if not, state that.
% @jovo: eq (15), i.e. S + W = T is well-known. Gabor and Rizzo wrote that
% in their paper, however, they didn't write a proof. They should have
% in some other paper, but I didn't bother looking and just did my own
% way.
% However, they mention 
% that this holds only for 0 < alpha < 2, with |x-y|^alpha, while
% here I did for a general semimetric. The PhD thesis do not 
% mention anywhere that
% minimizing W is equivalent to maximizing S, so I didn't see this written
% before. Anyway, this proposition is trivial, no big deal, but it's
% necessary to state it clearly for completeness

For a given $k$, the clustering problem amounts to
finding the best partition of the data by minimizing $W$.
% @gui: here and elsewhere is another good example.  why not just say "by minizing W"?
Notice that this is a hard clustering problem as partitions
are disjoint. The problem \eqref{eq:minimize} based on
energy statistics was already proposed in \cite{Kgroups}. However, it is
important to note that this is equivalent to maximizing \eqref{eq:between}
which is the test statistic for equality of distributions. In the following
we show what is the explicit optimization problem behind \eqref{eq:minimize}.

% @gui: "Let us" is used a bunch, though i think it never adds anything? i'd remove it everywhere.
% @jovo: will remove them
We now formulate \eqref{eq:minimize} in the corresponding
RKHS. Based on \eqref{eq:kernel_semimetric} and \eqref{eq:gen_kernel}, 
assume that the kernel $\kk: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ 
generates $\rho$.  Define  the Gram matrix
\begin{equation}
\label{eq:kernel_matrix}
G \equiv \begin{pmatrix}
\kk(x_1,x_1) & \kk(x_1,x_2) & \dotsm & \kk(x_1,x_n) \\
\kk(x_2,x_1) & \kk(x_2,x_2) & \dotsm & \kk(x_2,x_n) \\
\vdots & \vdots & \ddots  & \vdots \\
\kk(x_n,x_1) & \kk(x_n,x_2) & \dotsm & \kk(x_n,x_n) 
\end{pmatrix} .
\end{equation}
Let $Z \in \{ 0,1 \}^{n\times k}$ be the label matrix, 
with only one nonvanishing entry per row, 
indicating to which cluster (column)
each point (row) belongs to. This matrix satisfies
$Z^\top Z = D$ where the diagonal matrix 
$D = \diag( n_1,\dotsc, n_k )$  contains
the number of points in each cluster. We also define the rescaled
matrix  $Y \equiv Z D^{-1/2}$. In component form they are given by
\begin{equation}
\label{eq:label_matrix}
Z_{ij} \equiv \begin{cases}
1 & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} \qquad
\Zt_{ij} \equiv \begin{cases}
\tfrac{1}{\sqrt{n_j}} & \mbox{if $x_i \in \C_j$ } \\
0 & \mbox{otherwise}
\end{cases} .
\end{equation}
Throughout the paper, we use the notation $M_{i\bullet}$ to denote
the $i$th row of a matrix $M$, and $M_{\bullet j}$ denotes its $j$th column.
Our next result shows that the optimization problem \eqref{eq:minimize}
is NP-hard since
it is a quadratically constrained quadratic program (QCQP).
